Demystifying the Computer Science Ph.D. Admission in the US

### 2023-06-06 phd-cs-us.tex
updates

\documentclass[10pt]{article}
\usepackage[margin=1.1in]{geometry}
%\usepackage[small,compact]{titlesec} %very powerful
\usepackage{tcolorbox}
\usepackage{enumitem}
\usepackage[T1]{fontenc}
\usepackage{cite}
%\usepackage{caption}
%\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=blue,
    urlcolor=blue,
}
\usepackage{booktabs}

@@

\newcommand{\mycomment}[3][\color{black}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science Ph.D. Admission in the US}
\date{}
\author{ThanhVu (Vu) Nguyen, George Mason University}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

\begin{abstract}
Having been involved in PhD admissions for many years and after
numerous interaction with  students (especially Vietnamese), I've
realized that \emph{international} students lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

So I want to share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While I wrote this document for CS applicants, it should be applicable to various STEM fields.
Moreover, while many examples given are for GMU, which is 33rd on CSRankings (June 2016), the writing should be generalize to most other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US refers to a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools might be very extreme, e.g., see the admission process at \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{CMU}).

@@

First, I want to emphasize that PhD students in Computer
Science \emph{do not} need to worry about funding, especially at R1
universities in the US. If you are admitted, you will almost certainly
receive \emph{full funding} to support your study, including tuition,
health insurance, and 9-month stipend. Moreover, depending on the university,
you may even receive additional benefits like summer salary. Note that
funding is typically more available for PhD students than 
Masters. % [[#funding][Read more about funding here]]

Second, I believe that applying to a good US university \emph{should not} be any
harder than at good schools in other countries. If you think you have a
chance at schools in other countries, e.g., South Korea, Singapore, Germany, UK, Japan and Australia, then you surely have a chance in
the US as well.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: One of the reasons I create this post is that several of my colleagues are interested in 
recruiting Vietnamese students and were surprised when seeing very few Vietnamese applications (compared to other countries). In general the number of
PhD applications from Vietnam to US universities is very low and  more would be very welcomed. 
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: For me personally, it's better to have a good personalized
letter about your own research ability from someone who is less
well-known than a generic/weak letter from a well-known person.
\end{tcolorbox}

@@

This can be an issue for students in many top international universities where the competition is so high that very good students students can still have low ranking (and be overlooked by Admission committee).
So what to do with this? well, same as \hyperref[sec:your-school]{before}, e.g., put a note about this in your statement.

Note that while having good grades at unknown school might not help,
having very bad grades will be \red{red flag} (unless your LORs or
statements give proper explanation). This is especially true if you
have bad grades in relevant courses.

@@

You also do need to
take standardized English test. Just do well enough to pass minimum
requirement set by the university.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: The minimum for GMU CS (being above this might not mean much, but below is a \red{red flag}).
\begin{itemize}
\item GPA: $\ge 3.0$ in your undergrad (but as mentioned, we also consider the rank/prestige of your school)
\item GRE: not required (most CS graduate programs no longer require GRE, though it can help boost your profile)
%    \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
\item English requirement tests (one of the below)
  \begin{itemize}    
  \item TOEF: 88 pts in total AND $\ge 20$ points in each subsection OR
  \item IELTS: $\ge 6.5$ OR
  \item DuoLingo Graduate English: $\ge 120$ OR 
  \item Pearson Test of Academic English: $\ge 67$
  \end{itemize}  
\end{itemize}
\end{tcolorbox}

\subsection{CV/Resume}
This should be a summary of the accomplishments of the applicant.  I would quickly scan this document to see if something stand out (e.g., Publications, Programming Competition Awards, Teaching Experiences).

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
At GMU, we are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some rare cases a faculty interviews a candidate that they see potentials and want argue for admission, i.e., without the interview, that candidate is definitely rejected. In any case, getting interview means you have a very good chance of being admitted.
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
At GMU, TA and RA have similar benefits in tuition waiving and insurance.  For stipend, depending on the college and department, a 9-month graduate assistant stipend is set.  TA and RA will usually be that amount (TA will definitely be that, RA might fluctuate a bit depending on the stage of the student and the prof.). 
\end{tcolorbox}

@@

\subsection{Current 2023 Rankings of CS PhD programs in the U.S}
  Below is the 2023 Rankings from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system that is based on top CS conferences, as of April 2023.

>>>>

\documentclass[11pt]{article}
\usepackage[margin=1.1in]{geometry}
%\usepackage[small,compact]{titlesec} %very powerful
\usepackage{tcolorbox}
\usepackage{enumitem}
\usepackage[T1]{fontenc}
\usepackage{cite}
%\usepackage{caption}
%\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=blue,
    urlcolor=blue,
}
\usepackage{booktabs}

@@

\newcommand{\mycomment}[3][\color{black}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science Ph.D. Admission in the US}
\date{\today}
\author{ThanhVu (Vu) Nguyen, George Mason University}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

\begin{abstract}
Having been involved in PhD admissions for many years and after
numerous interaction with  students, I've
realized that \emph{international} students, especially Vietnamese, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

So I want to share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While I wrote this document for Vietnamese students interested in CS, it should be applicable to students in various countries interested in STEM.
Moreover, while many examples given are for GMU, which is 33rd on CSRankings (see \S\ref{sec:ranking}), the writing should be generalize to most other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US refers to a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools might be very extreme, e.g., see the admission process at \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{CMU}).

@@

First, I want to emphasize that PhD students in Computer
Science \emph{do not} need to worry about funding, especially at R1
universities in the US. If you are admitted, you will almost certainly
receive \emph{full funding} to support your study, including tuition,
health insurance, and 9-month stipend. Moreover, depending on the university,
you may even receive additional benefits like summer salary. Note that
funding is typically more available for PhD students than 
Masters and especially undergraduate studies, which typically have no funding at all. \S\ref{sec:funding} provides more details on funding.

Second, I believe that applying to a good US university \emph{should not} be any
harder than at schools in other countries. If you think you have a
chance at schools in other countries, e.g., South Korea, Singapore, Germany, UK, Japan and Australia, then you will surely have a chance in the US as well.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: One of the reasons I create this post is that several of my colleagues are interested in 
recruiting Vietnamese students and were surprised when seeing very few applications in Vietnam (compared to other countries). In general the number of
PhD applications from Vietnam to US universities is very low and  more would be very welcomed. 
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: It's better to have a good personalized
letter about your own research ability from someone who is less
well-known than a generic/weak letter from a well-known person.

Vietnamese students sometimes go through great length just to get letters from well-known senior professors in their school, but the letters are very generic and weak and therefore carry little value, in fact, red flags. Moreover, a top professor in Vietnamese might not be well-known to US faculty (see more details in \S\ref{sec:your-school}). So save the trouble and just get letters from \emph{any} professors who knows you well and can write a good letter about your research ability.
\end{tcolorbox}

@@

This can be an issue for students in many top international universities where the competition is so high that very good students students can still have low ranking (and be overlooked by Admission committee).
So what to do with this? well, same as \hyperref[sec:your-school]{before}, e.g., put a note about this in your statement and ask your LoR writers to talk about it.

Note that while having good grades at unknown school might not help,
having very bad grades will be \red{red flag} (unless your LORs or
statements give proper explanation). This is especially true if you
have bad grades in relevant, e.g., CS and Math, courses.

@@

\paragraph{English Test} Unless you're from certain countries (e.g., English speaking ones), you will need to
take standardized English test. Just do well enough to pass minimum requirement set by the university, which nowadays has many options for you to choose from.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: The minimum for GMU (being above this might not mean much, but below is a \red{red flag}).
\begin{itemize}
\item GPA: $\ge 3.0$ in your undergrad (but as mentioned, we also consider the rank/prestige of your school)
\item GRE: not required, though it can help boost your profile
%    \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
\item English requirement tests (one of the below)
  \begin{itemize}    
  \item TOEF: 88 pts in total AND $\ge 20$ points in each subsection OR
  \item IELTS: $\ge 6.5$ OR
  \item DuoLingo Graduate English: $\ge 120$ OR 
  \item Pearson Test of Academic English: $\ge 67$
  \end{itemize}  
\end{itemize}
\end{tcolorbox}

\subsection{CV/Resume}
This should be a summary of the accomplishments of the applicant.  It should allow the reviewers to quickly scan to identify standout achievements (e.g., Publications, Programming Competition Awards, Teaching Experience).

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: At GMU, we are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some rare cases a faculty interviews a candidate that they see potentials and want argue for admission, i.e., without the interview, that candidate is definitely rejected. In any case, getting interview means you have a very good chance of being admitted.
\end{tcolorbox}

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: TA and RA at GMU have similar benefits in tuition waiving and insurance.  For stipend, depending on the college and department, a 9-month graduate assistant stipend is set.  TA and RA will usually be that amount (TA will definitely be that, RA might fluctuate a bit depending on the stage of the student and the prof.). 
\end{tcolorbox}

@@

\subsection{Current 2023 Rankings of CS PhD programs in the U.S}\label{sec:ranking}
  Below is the 2023 Rankings from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system that is based on top CS conferences, as of April 2023.
